---
title: "Project"
author: "Lucas Villalba"
date: "2024-01-13"
output: html_document
---

# Capstone Project
## Part of Google Data Analytics profesional certificate

Working on how the Garmin personal database(could be any other brand) can help us throughout exams at the third level education with a healthy body equals to a healthy mind,

The dataset the physical activities, habits, and routines of a male age 39 ~ 42 years old, from 2018 to 2022.

How a mature student can cope throughout exams at third level education and success with a good health and fitness, 'mens sana in corpore sano'

Also, it enhances our readiness for a future successful professional life after finishing our studies.


* The methodology going through the Ask, Prepare, Process, Analyze, and Share phases of the data analysis process.

## ASK

# Five questions:

1. What type of company does your client represent, and what are they asking you to accomplish?\  

Personal interest in how I can cope with stress during exams

2. What are the key factors involved in the business task you are investigating?\  

Tracking my physical activity during the 4 (four) years, the same time period which took to completed my course. 

3. What type of data will be appropriate for your analysis?\  

All my physical activities during the time from 2018 to 2022.

4. Where will you obtain that data?\  

The data obtained from Garmin personal fitness database, downloaded via Garmin website from 2018 to 2022.

5. Who is your audience, and what materials will help you present to them effectively?\  

My audience a mature male group range age from 38 years old to 42 years old, who are attending a full time course, working part time and in a professional growth path.


*Stating a business task clearly.*\  

**How a healthy physical activity routine helps us to cope with stress during exams.**
```{r, setting un CRAN environment}
options(repos = c(CRAN = "https://cran.rstudio.com/"))
install.packages("tidyverse")

```

* Importing data from a real dataset.
```{r, importing data}
activities_df <-
  read.csv("~/cloud/project/Activities.csv")
str(activities_df)
```

* Documenting any data cleaning that you perform on the data set.
```{r, data cleaning}
install.packages("tidyverse")
install.packages("skimr")
install.packages("janitor")
library(tidyverse)
library(skimr)
library(janitor)
```

* Quick visualization of the raw data.
```{r, creating a visualization }
ggplot(data = activities_df) +
  geom_point(mapping = aes(x = Activity.Type, y = Calories, color = Activity.Type)) +
  labs(title = "All the activities by Calories") +
  theme(axis.text.x = element_text(angle = 45))
```

# Prepare

* Where is your data located?  The Data is located in a sub-folder no related to the project, for this reason is good to save the data in the following path "R/Coursera/Project" this is part of good habits data structure

```{r, data relocate in Project subfolder}
write.csv(activities_df, file = "~/cloud/project/activities1.csv", na = "NA")

```

* How is the data organized?  Data frame is organized in rows and columns, with 854 observations (rows) and 42 variables (columns), each observation is a physical activity with variables like:
```{r, variables}
colnames(activities_df)
```


```{r, data overview}
skim_without_charts(activities_df)
```


* Are there issues with bias or credibility in this data? Does your data ROCCC?
 The data has tracked and collected via smart watch Garmin model Forerunner 25 and forerunner 35.
 The data is Reliable as the device to collect the data uses a GPS and sensors for standard accuracy level.
 The data is Original direct from personal Garmin database.
 The data is comprehensive, physical activities since 2018 to 2022, same time the subject was attending 4 years  full time undergraduate course.
 Current. the data is actual for the time period to perform the analysis.
 Cited, the data is from a trusted private service with customer support and trusted system.

* How are you addressing licensing, privacy, security, and accessibility?
The data frame is a personal data, saved in a sub folder without security measures, free access.

* How did you verify the data's integrity?
The data has been downloaded from a Garmin website, access by username and password to download it.
The data has been collected via Garmin smart watch since 2018 to 2022 of an male adult person age 40yrs old, same time the subject was attending to undergraduate course 4 years full time.

* How does it help you answer your question?
Yes, the data as a track of all the physical activities of the subject from 2018 to 2022, same time the subject was attending college.

* Are there any problems with the data?
No problem with the data.

Key tasks

* Download data and store it appropriately.
* Identify how it’s organized.
* Sort and filter the data.
* Determine the credibility of the data.

Deliverable

* A description of all data sources used

# Process

* What tools are you choosing and why?  I am choosing RStudio RStudio
2023.12.0 Build 369
© 2009-2023 Posit Software, PBC
I find it more efficient for data visualization, cleaning and data wrangling, all with a same document in RMarkdown. 

* Have you ensured your data’s integrity?  The source of the data is trusted.

* What steps have you taken to ensure that your data is clean?  Data cleaning steps:
```{r, Data cleaning detailed process}

# Select the variables for the analysis.
trimmed_activities_2018_2022<-activities_df %>%
  select(Activity.Type, Date, Calories, Avg.HR, Avg.Stress)

# Check the data type of the variables.
str(trimmed_activities_2018_2022$Date)

# Change the variable Date (character type) to date(date type) .
trimmed_activities_2018_2022$date <- as.Date(trimmed_activities_2018_2022$Date)

# Change the variable Calories(character type) to calories(numeric type).
trimmed_activities_2018_2022$calories <- as.numeric(trimmed_activities_2018_2022$Calories)

# Change the variable Avg.HR (character type) to avg_hr (numeric type).
trimmed_activities_2018_2022$avg_hr <- as.numeric(trimmed_activities_2018_2022$Avg.HR)

# Change the variable Avg.Stress (character type) to avg_stress (numeric type).
trimmed_activities_2018_2022$avg_stress <- as.numeric(trimmed_activities_2018_2022$Avg.Stress)

# Change the variable name Activity.Type to act_type and delet from the data frame.
trimmed_activities_2018_2022 <- trimmed_activities_2018_2022 %>%
  mutate(act_type = Activity.Type) %>%
  select(-Activity.Type)

# finally select the new order for the variables for analysis.
act_2018_2022 <- trimmed_activities_2018_2022 %>%
  select(act_type, date, avg_stress, avg_hr, calories)
act_2018_2022$date <- as.Date(act_2018_2022$date)
class(act_2018_2022$date)
```

* How can you verify that your data is clean and ready to analyze?  Check data structure.
```{r check the structure to finish step of the data wrangling, warning=FALSE}
# Overview of the data frame
str(act_2018_2022)
colnames(act_2018_2022)
skim_without_charts(act_2018_2022)

# dealing with missing values
# the device(forerunner 35) does not read stress so for three first year was no data, so we can see a small data set with out NA or missing values, decided to keep the NA and using calories variable.
drop_na(act_2018_2022)

# Data visualization of the data cleaning results
ggplot(data = act_2018_2022) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by Calories") +
  theme(axis.text.x = element_text(angle = 45))

# Data manipulation, group the data by academic calendar year as following:

# Calendar year from Sept 2018 to Jun 2019
act_2018_2019 <- act_2018_2022[act_2018_2022$date > "2018-09-01" & act_2018_2022$date < "2019-06-30",]

# Calendar year from Sept 2019 to Jun 2020
act_2019_2020 <- act_2018_2022[act_2018_2022$date > "2019-09-01" & act_2018_2022$date < "2020-06-30",]

# Calendar year from Sept 2020 to Jun 2021
act_2020_2021 <- act_2018_2022[act_2018_2022$date > "2020-09-01" & act_2018_2022$date < "2021-06-30",]

# Calendar year from Sept 2021 to Jun 2022
act_2021_2022 <- act_2018_2022[act_2018_2022$date > "2021-09-01" & act_2018_2022$date < "2022-06-30",]

# Visualizations by academic year
# Activities 1st year 2018 2019
ggplot(data = act_2018_2019) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by date in the 1st academic year(2018 2019)") +
  theme(axis.text.x = element_text(angle = 45))

# Activities 2nd year 2019 2020
ggplot(data = act_2019_2020) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by date in the 2nd academic year(2019 2020)")+
  theme(axis.text.x = element_text(angle = 45))

# Activities 3rd year 2020 2021
ggplot(data = act_2020_2021) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by date in the 3rd academic year(2020 2021)") +
  theme(axis.text.x = element_text(angle = 45))

# Activities 4th year 2021 2022
ggplot(data = act_2021_2022) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by date in the 4th academic year(2021 2022)") +
  theme(axis.text.x = element_text(angle = 45))

```


* Have you documented your cleaning process so you can review and share those results?
The data still have 35 missing values, but this do not affect the analysis due the calories variable is taking in account for this.
Key tasks
* Check the data for errors.
* Choose your tools.
* Transform the data so you can work with it effectively.
* Document the cleaning process.

Deliverable
Documentation of any cleaning or manipulation of data

# Analyze

* Analyzing the data with visualizations year by year.

* Creating data visualizations from your analysis.
```{r, Analysing the data}
# Calculate the start date and the finish date
min_d_2018<-min(act_2018_2019$date)
max_d_2019<-max(act_2018_2019$date)

# Visualization of the year with date and exams period
p<-ggplot(data = act_2018_2019) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by date in the 1st year", caption=paste0("Data from: ", min_d_2018, " to ", max_d_2019), x = "Date", y = "Activity type") +
  theme(axis.text.x = element_text(angle = 45))
  
# adding the exams period
p + annotate(geom = "text", x = as.Date("2019-01-11"), y =(act_2018_2019$act_type), label = "Exams period", vjust = -2, color = "orange")

```


## 1st year
For the first year, the exam period was predicted in January, no repeating exams period as was not re-sitting.
**"Exams period"** shows the lack of physical activity and this trigger stress as new physical activities were in place after the exams period, as *walking*.

```{r}
# Calculate the start date and the finish date
min_d_2019<-min(act_2019_2020$date)
max_d_2020<-max(act_2019_2020$date)

# Visualization of the year with date and exams period
p<-ggplot(data = act_2019_2020) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by date in the 2nd year", caption=paste0("Data from: ", min_d_2019, " to ", max_d_2020), x = "Date", y = "Activity type") +
  theme(axis.text.x = element_text(angle = 45))
  
# adding the exams period
p + annotate(geom = "text", x = as.Date("2020-01-02"), y =(act_2019_2020$act_type), label = "Exams period", vjust = -2, color = "orange")
```


## 2nd year
For the second year, the exam period was predicted in January, no repeating exams period as was not re-sitting.
**"Exams period"** shows more physical activity compare with previous year, more calories as more physical activities were in place during the exams period *Calories 750*.



```{r}
# Calculate the start date and the finish date
min_d_2020<-min(act_2020_2021$date)
max_d_2021<-max(act_2020_2021$date)

# Visualization of the year with date and exams period
p<-ggplot(data = act_2020_2021) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by date in the 3rd year", caption=paste0("Data from: ", min_d_2020, " to ", max_d_2021), x = "Date", y = "Activity type") +
  theme(axis.text.x = element_text(angle = 45))
  
# adding the exams period
p + annotate(geom = "text", x = as.Date("2021-01-02"), y =(act_2020_2021$act_type), label = "Exams period", vjust = -2, color = "orange")
```


## 3rd year
For the third year, the exam period was predicted in January, no repeating exams period as was not re-sitting.
**"Exams period"** shows the continuity of physical activity, this keep calories consumption levels, and new physical activities were in place at the end of exams period, as *Street Running* and *Strength training*.

```{r}
# Calculate the start date and the finish date
min_d_2021<-min(act_2021_2022$date)
max_d_2022<-max(act_2021_2022$date)

# Visualization of the year with date and exams period
p<-ggplot(data = act_2021_2022) +
  geom_point(mapping = aes(x = date, y = act_type, color = act_type, size = calories)) +
  labs(title = "Activities by date in the 4th year", caption=paste0("Data from: ", min_d_2021, " to ", max_d_2022), x = "Date", y = "Activity type") +
  theme(axis.text.x = element_text(angle = 45))
  
# adding the exams period
p + annotate(geom = "text", x = as.Date("2022-01-02"), y =(act_2021_2022$act_type), label = "Exams period", vjust = -2, color = "orange")
```


## 4th year
For the fourth year, the exam period was predicted in January, no repeating exams period as was not re-sitting.
**"Exams period"** shows a decrease of physical activity, a high stress post exams period and a new physical activities added in place after the exams period, as *Yoga* and *Pilates*.

## Final Conclusion:
For the final result of the analysis stated the physical activities with high intensity(hart rate) and workout(calories) as *Running*, *Cycling*, *Cardio* activities, which put the subject in a **high stress**, on the other hand, the activities post high stress period are more accurate in relation to **cope with stress** as *Walking*, *Yoga* and *Pilates*.


* Summarizing key findings from your analysis.

* Documenting your conclusions and recommendations.

* Creating and publishing your case study.


Deliverable:
1. A clear statement of the business task you have selected to investigate
2. A description of all data sources used
3. Documentation of any cleaning or manipulation of data
4. A summary of your analysis
5. Supporting visualizations and key findings
6. Based on what you discover, a list of additional deliverables you think would be helpful to include for further exploration
7. Your top high-level insights based on your analysis
